draft 1

### 2016-07-15 finalwriteup/draft.rmd
finished editing method

The literature pertaining to the aim of this research project, cost prediction using internal consulting CRM data, has been extensively reviewed. The problem has been addressed in studies beforehand which have generally been limited to prediction of construction costs for buildings/infrastructure and effort estimation for IT projects. The prediction models consisted of Case Based Reasoning, Linear Regression, Neural Networks and in one or two instances ensemble trees or SVM. This leaves a gap in effort estimation for consulting companies in the construction industry in combination with trialing a broader range of prediction methods.

@@

The aim of this project is to determine whether the profitability of projects for consulting businesses can be predicted using statistical and machine learning techniques. As stated in the literature review, a couple of gaps exist in this research space including working with internal CRM data from a consulting company in the construction industry, trialing ensemble tree methods against the more popular regression and Neural Networks, and analysing the impact model results could have on business decisions. This section will explain the chosen methodology for fulfilling the research aim using a case study engineering consulting company that charges for their time spent on project delivery.

@@

An assessment of the literature revealed that the capacity of internal timesheet data (extracted from a company's CRM software) to predict project profitability has not yet been tested in the context of consulting companies in the construction industry. Project data can be extracted from the CRM, interviewing project team members, emails, accounting records, and project drawings and calculations. As the internal CRM data in this case study houses a detailed account of time spent on each project (timesheet data) along with client information and invoicing records for the preceding twelve years, it was clear that this source of data was not only the most accessible and suitable for statistical analysis, but had the potential to reveal information not typically investigated by the company.

The complete twelve years of project timesheets and invoices was extracted via the CRM software interface, which described 4169 projects. The projects varied from total invoiced amounts of \$500 to over \$1,000,000 and were divided between four internal disciplines in the company. During data extraction, it was not clear whether the full twelve years were necessary and relevant to predicting future project profitability but this was determined later in the analysis.

Once the data was extracted from the CRM, an employee from the case study company performed the task of de-identification. This is an important step before research commences as the privacy of project employees, clients, and project names must be removed for ethical reasons. In the case of employee names, professional titles replaced the names, and client codes replaced client names. Project names and addresses were removed, and each project was identified by a number. The resulting de-identified datasets were three .csv files of different data structures which are described in the next section.

## Cleaning and variable engineering

The data contained a rich source of project information and a lengthy process of cleaning the data followed by engineering potentially useful variables was then required. The three initial datasets described:

* Invoicing records: dated invoiced amounts and sizes for each project and whether the invoice had been paid or written off

@@

| Date    | Hours | Charge Amount | Cost Amount | Created By | Discipline | Job No. | Reconciled | Scope       |
|---------|-------|---------------|-------------|------------|------------|---------|------------|-------------|
| 1.2.300 | 4     | 640           | 400         | P12        | Civil      | 1.2.300 | Yes        | Concept     |
| 1.2.400 | 4.5   | 720           | 450         | P12        | Civil      | 1.2.300 | Yes        | Preliminary |
| 1.2.500 | 2     | 480           | 320         | P34        | Civil      | 1.2.300 | Yes        | Preliminary |
| 1.2.600 | 7     | 1120          | 700         | P12        | Civil      | 1.2.300 | Yes        | Preliminary |
| 1.2.700 | 1     | 160           | 100         | P12        | Civil      | 1.2.300 | Yes        | Preliminary |
| 1.2.800 | 3     | 390           | 240         | P3         | Structural | 1.2.400 | Yes        | Site        |
| ...     | ...   | ...           | ...         | ...        | ...        |         | ...        | ...         |\
\\
\

@@

These three sources needed to be compiled into a single dataset that detailed one project per row as the model was to predict the overall profitability of a project. Before compilation could begin however, thorough cleaning was required where the data was plotted and statistically assessed so that outliers could be visually or statistically discovered. Outliers were then investigated for data entry errors. Many such errors were encountered. For example, a common glitch in the data collection process translated a user entering a 123 kilometer drive in a car as 123 hours. Or an expense printing claim of $9.50 would be translated to 9.5 hours spent on that job. Once these errors were detected, they were discussed with the company directors and appropriate corrective action was taken. Mistakes in data entry were often able to be converted into correct data by notes that a user had entered into the system. It is possible that some data entry errors were not detected and this is a problem that must be dealt with in implementation of the model as well.

@@

Although dozens of variables were available from the initial dataset, it was advantageous to engineer further descriptive variables to trial in the models. This was particularly true for the invoicing dataset and timesheet dataset where a project could have thousands of rows of relevant data that needed to be converted into a single row per project. All variables would eventually be tested for variable importance and predictive power with respect to project profitability. Examples of these engineered variables include:

@@

* Percent of hours performed by each professional role over the course of a project
* Timespan of the entered project hours
* Percent of hours performed by 'professional' employees as opposed to 'technical' employees
* Position of the employee who completed the most hours on each project
* Percent of hours done by the majority contributor to a project
* Total cost of employee hours per project
* Total cost of external subcontractors or disbursements per project
* Total number of users that entered hours on each project
* Mean number of hours per day entered on a project
* Number of disciplines active in a project 
\ 
\

@@

* Total amount invoiced and remunerated per project
* Mean invoice size per project (for example if a project had four invoices totaling \$10,000, the mean invoice size would be \$2,500)
* Mean invoice per client, i.e. the average invoice size across all invoices sent to a client. This gives an indication of whether the client generally does fast paced big jobs which would result in large monthly invoices for example or small jobs with small monthly invoiced amounts.
* Invoice frequency per project. This indicates how drawn out a project was.
* Client invoice frequency which gives an indication of how much work the company does for that client. 
\ 
\

@@

* text analysis of project descriptions detected a list of key words that could classify projects into 16 categories. This key word analysis was done in conjunction with a company employee and the resulting classifications were reviewed by the employee to ensure accuracy.
* number of projects completed with each client and client contact

With the engineered variables, all three data sources could be represented in the format of a single row per project. They were hten combined and further variables were engineered using combinations of variables across the three data sets:

@@

Besides the engineered variables, additional variables included in the project summary dataset were client industry, internal company discipline, job description, and post code. Once numerous potentially important variables were engineered, the most important ones could then be narrowed down. This practice improves the accuracy of a model because unnecessary or irrelevant variables add noise to the prediction of target values. Variable selection also increases computational efficiency by reducing the number of calculations and enhances understanding of the prediction structure [@Weisberg2005].

@@

Once variables were engineered, the value each variable could contribute to the predictive model was assessed. As previously stated, variable selection can enhance model accuracy by reducing noise from irrelevant variables and increase simplicity which in turn improves interpretability for the user. To assess variable importance, models predicting 'return per dollar' were built using all available explanatory variables. The models then output which variables contributed significantly or most improved results. Different types of predictive models have different methods of calculating variable importance and since it was not known which model would best predict return per dollar, a few models' variable outputs were reviewed. These included ANOVA Linear Regression and two ensemble tree methods. Before models assessing variable importance were built, outlier cases were deleted to prevent distortion of valuable explanatory variables.

@@

Case study projects with extreme 'return per dollar' values were likely the result of special scenarios which the model is not intended to predict. For example, many projects appeared wildly profitable but on closer inspection these were all very small jobs. After discussion with the case study company, it was concluded that for those outliers, the small jobs most likely required so little time that busy employees did not bother writing down their hours for that job. Outliers were initially defined as return per dollar values sitting outside 1.5x the interquartile range (IQR) from the upper and lower quartiles. This rule was initially practised by Tukey, the inventor of the box plot [@Tukey1977]. All outlier projects determined by this method were reviewed with an employee from the case study company and based on their domain knowledge and assessment of the outlier cases, reasonable cut-off values for the range of return per dollar values were determined. In summary, projects with 'return per dollar' values greater than 3 or below -2 were removed.

@@

Three methods which output variable importance were trialed and compared to test whether they highlighted different variables as valuable. This was anticipated as each method models data using distinct theoretical foundations. The tested methods were ANOVA Linear Regression, conditional inference forests (cForests), and Random Forests. It was expected that Random Forests would favour variables that had more categories, a well-known bias; which is also a bias that cForests have overcome [@Strobl2007]. ANOVA is based on linear theory which contrasts conditional forest and Random Forest's free-form structure. Therefore, it is beneficial to compare the methods' assessments of variable importance.

When running variable importance tests, the dependent variable must be clearly defined. For this problem, the response variable could be continuous or categorical i.e. *how* profitable a project was (continuous) or whether a project was profitable or not (binary classification). The continuous response variable was chosen for the variable importance models as the *degree* of profitability also gives the binary classification (profitable or unprofitable).

#### ANOVA

ANOVA is closely related to linear fit models but incorporates the analysis of differences in group means (i.e. categorical variables) [@Lunney1970]. The method is simple and fast but assumes variables are normally distributed and linearly related. The variables must therefore be normalised before analysis to achieve a statistical distribution as close to Gaussian as possible. A disadvantage of this method is that linear relationships and normally distributed variables do not necessarily represent real world data accurately [@Breiman2001b].

In order to compare variable importance, two types of output from the model are assessed which include the linear coefficients for each variable and the p-values for the coefficients [@Markham2016]. The magnitude of a coefficient does not indicate importance relative to the other variables because its value is directly reflects the values within the variable. The sign of the coefficient however indicates whether the variable is directly or inversely related to the response variable. A p-values for each variable coefficient is output as the result of a significance test for whether the coefficient equals zero. Therefore, if the null hypothesis is rejected (p value < 0.05), it indicates that changes in the explanatory variable are related to changes in the response variable. P-values are a good indication of whether variables have a relationship with the target variable, however they do not necessarily rank importance. If a variable is measured more precisely, it will have a smaller p-value, whereas a variable measured roughly will have a higher p-value. This does not necessarily make the more precisely measure variable is more important than the other. It is also important to keep in mind that, with a dozen variables, chance alone can produce a variable with a p-value < 0.001 7% of the time [@Rice1989].

@@

#### Random Forest

The Random Forest algorithm can produce a permutation variable importance for each variable as part of its output. Variable importance is represented as a score, which is calculated by performing permutations on each variable in the test data for each tree and re-running the test data through the trees. More specifically, consider a single variable. The values of this variable in the out-of-bag sample (the test data that the Random Forest algorithm partitions before building each tree) are randomly permuted (reordered) for each tree. The purpose of permuting is to mimic the absence of that variable. The trees in the Random Forest are then run again. Variable importance can now be derived by comparing the results of the forest where the covariate has been permuted to the original Random Forest. If the accuracy of the permuted forest is much less than the original forest, then that variable was important. Permutation importance scores are calculated as the mean decrease in accuracy over all trees for that permuted variable in the Random Forest [@Breiman2005].

The Random Forest variable importance provides valuable insight as
it addresses the impact of each predictor variable individually as well as in multivariate interactions with other predictor variables [@Strobl2007]. However, a severe disadvantage of Random Forest variable importance is that it is not reliable for variable selection where explanatory variables vary in their scale of measurement or their count of categories. Also, the importance of correlated predictors is overestimated and the algorithm tends to favour variables that have many possible splits or many missing values [@Strobl2007]. The intuition behind this is that if a variable has more points to split the data, there are more opportunities for it to split the response variable favourably, however this does not necessarily indicate the variable impacts the response variable most. Overall, the Random Forest permutation importance is useful as it is able to compare variables in a machine learning environment and can include multivariate interactions, but the current Random Forest importance measure overestimates the importance of variables with many categories and some numeric variables.

#### Cforests

Cforests are an alternative ensemble tree method to Random Forests that overcome the aforementioned shortfalls of Random Forest variable importance measures. It is built from cTrees which are decision trees based on a conditional inference framework. The key difference between cTrees and standard decision trees is that a significance test is used for splitting instead of a purity measure such as the Gini coefficient, which is centered around information gain.

The procedure behind the significance test at each split in a cTree is as follows. To determine which explanatory variable should be used at each split, each variable is permuted in every possible way, and a correlation value is calculated between the tested variable and response variable, for each permutation. The unchanged variable correlation is then compared with the correlation values for all permutations of that variable. From this, a p-value for the true correlation value compared to the permuted correlation values can be calculated. The predictor variable with the lowest p-value is then selected as the splitting variable [@Hothorn2006]. Using an ensemble forest of these trees, variable importance is then calculated in the same fashion as Random Forests, via permuting a variable, re-running the forest and comparing the decrease in response variable accuracy. Several sources recommend cForests variable importance methods because its statistical p-value tests at each split removes the bias present in other tree ensemble variable importance outcomes such as Random Forests (as discussed) as well as Boosted Trees [@Strobl2007][@Strobl2009][@Hothorn2006].

@@

Limiting the predictive model to a concise set of meaningful variables reduces noise and improves predictions. Less variables means that a simpler model is being used for prediction which in turn is easier for stakeholders to understand [@Weisberg2005]. For these reasons, a subset of the most important variables were chosen before modelling which entailed first eliminating outliers, then comparing important variables from ANOVA as well as Random Forests and cForests. It can be inferred from the literature that cForests would suit the case study data best as it is unbiased and not limited to linear theory [@Strobl2007].

@@

Previous studies predicting project profitability in software and construction projects were predominantly limited to case-based reasoning, regression and Neural Networks. There is a need to test other sophisticated machine learning algorithms that are as powerful as Neural Networks but provide insight into the reasoning behind predictions. Ensemble trees (Boosted Trees and Random Forests) and Bayesian Networks fit these criteria. Regression and Naive Bayes models were also included as simple baseline models. Complex models should be measured against simple models that can be built at a fraction of the computational cost. The complete list of models tested in this study is:

* Regression - baseline model
* Naive Bayes - baseline model
* Bayesian network
* Random forest
* Gradient Boosted Trees

Some models have limitations and required additional data processing steps such as normalising numeric variables (Bayesian Networks and Regression) and discretising continuous variables (Bayesian Networks). Discretising numeric variables was performed by generating a hierarchical dendrogram of each variable to visualise the clusters. Between four and six clusters were chosen and summarised to find the maximum and minimum values within each cluster. For example, the diagram below illustrates the hierarchical dendrogram for timespan with 6 clusters boxed.

@@

The number of clusters was determined both visually from the dendrogram and experimentally. Visually, the height of the dendrogram 'branches' indicates the change in 'tightness' of the data points to their cluster centroids as the number of clusters increases. Tightness in the case of Ward's method is the sum of squared distances of each data point to the centroid of its respective cluster (ESS). For example, the height of the top horizontal bar indicates the ESS for one cluster and the height of the second from top horizontal bar is the ESS for two clusters. The height of the vertical branch between the two bars indicates the improvement in ESS by adding a cluster [@Putler2012]. As more clusters are added, decreases in ESS are less significant and the aim is to choose the number of clusters at an optimal level in this process.  After initial comparisons of prediction methods had been trialed, the number of clusters for each discretised variable was re-assessed by trialing one more or one less cluster and comparing model output.

@@

@@

All methods except gradient Boosted Trees could not handle missing data. Therefore, preliminary runs of each method used subsets of the data which had complete data. It was possible that if missing data was imputed, predictions from the models that were limited by missing data could improve. Also, a complete data set allows for complete sets of predictions from each method, which can then be blended to further improve results. First, imputation methods were trialed against gradient Boosted Tree results since Boosted Trees can handle missing values and should perform equally well with imputed data.

The MICE Random Forest method was chosen because it has been proven to work well with complex data sets [@Shah2014]. It first performs a standard Random Forest imputation of missing values which are treated as initial 'place holders'. Random forest imputation is a complex process which works by first doing a rough imputation of the missing values. Then a ten-tree forest is run to calculate proximities between all cases. Proximities between two cases represent the number of trees in a forest where the two cases were assigned to the same end node in a tree. Once proximities are calculated, missing values for a variable (say variable m) are filled by an average of the other complete m values weighted by the proximity of the case representing each m value.

@@

To compare the models, root mean squared error (RMSE) statistic was used for regression models and the area under the receiver operating characteristic (ROC) curve (AUC) statistic was used to measure binary classification. For binary classification, AUC is a more useful statistic than classification accuracy when output is available as a probability which can be applied to the problem. This is pertinent when it is useful to rank the outcomes and trial probability thresholds that will optimise the rates of true positives and true negatives [@Huang2005]. Note the 'probability' outputs from the models are not true statistical probabilities, but a score the model has assigned to a case using its own measure of certainty.

An ROC graph visualises the curve from which an AUC score is calculated. Its two axes are the false positive rate (FPR) on the x-axis and true positive rate (TPR) on the y-axis. In this problem a 'positive' is a loss-making job so a true positive (TP) is a case when the model predicts a job will be loss making, and it did indeed lose money. A false positive (FP) is when the model predicts a job will be loss making but in reality it was profitable. These values are calculated as follows:

Only predictions made on the test data should be entered into the equations above, therefore after a model must be first created using a training set, then the test data is input. From the test data, the model outputs probability values between 0 and 1 for each case, which cannot be entered into the equations above. In order to classify each case in the test set as 0 or 1, a probability threshold must be chosen. That is, if a probability threshold is chosen to be 0.6, each case in the test set can then be classified as 0 or 1 depending on whether the probability calculated by the model is greater or less than 0.6. Once the classifications have been made, the TPR and FPR can be calculated for that model and threshold. This is plotted as a single point on the ROC curve, say point C. Refer the diagram below:

>>>>

The literature pertaining to the aim of this research project, cost prediction using internal consulting CRM data, has been extensively reviewed. Previous studies addressing the problem have generally been limited to prediction of construction costs for buildings/infrastructure and effort estimation for IT projects. The prediction models consisted of Case Based Reasoning, Linear Regression, Neural Networks and in one or two instances ensemble trees or SVM. This leaves a gap in effort estimation for consulting companies in the construction industry in combination with trialing a broader range of prediction methods.

@@

This section outlines the chosen methodology for fulfilling the research aim - predicting project profitability using internal company data and testing how much it improves the business' bottom line. The process of how the data was collected, cleaned, and how the predictive models were progressively built and tested is explained along with why the steps were taken. This section should enable an experienced practioner to replicate the study.

@@

An assessment of the literature revealed that the capacity of internal timesheet data (extracted from a company's CRM software) to predict project profitability has not yet been tested in the context of consulting companies in the construction industry. Alternate sources of project data were considered including interviewing project team members, emails, accounting records, and project drawings and calculations. However the CRM data from this case study could be extracted in a spreadsheet format and housed a detailed account of time spent on each project (timesheet data) along with client information and invoicing records for the preceding twelve years. This source of data was not only the most accessible and suitable for statistical analysis, but had the potential to reveal information not typically investigated by the company.

The twelve years of project data was extracted via the CRM software interface, which described 4169 projects. Projects varied from total invoiced amounts of \$500 to over \$1,000,000 and were divided between four internal company disciplines. During data extraction, it was not clear whether the full twelve years were relevant to predicting future project profitability but this was determined later in the analysis.

Once the data was extracted from the CRM, an employee from the case study company performed the task of de-identification. This is an important step before research commences as the privacy of project employees, clients, and project names must be removed for ethical reasons. In the case of employee names, professional titles replaced the names, and client codes replaced client names. Project names and addresses were removed, and each project was identified by a number. The resulting de-identified data consisted of three .csv files of different data structures which are described in the next section.

## Cleaning and Variable Engineering

The data contained a rich source of project information and a lengthy process of cleaning the data was required. Then, potentially useful variables were engineered which could be tested alongside original variables. The three initial datasets described:

* Invoicing records: dated invoiced amounts for each project and whether the invoice had been paid or alternatively, written off.

@@

| Date    | Hours | Charge Amount (\$)| Cost Amount (\$)| Created By | Discipline | Job No. | Reconciled | Scope       |
|---------|-------|---------------|-------------|------------|------------|---------|------------|-------------|
| 2014/09/01 | 4     | 640           | 400         | P12        | Civil      | 1.2.300 | Yes        | Concept     |
| 2014/09/03 | 4.5   | 720           | 450         | P12        | Civil      | 1.2.300 | Yes        | Preliminary |
| 2014/09/03 | 2     | 480           | 320         | P34        | Civil      | 1.2.300 | Yes        | Preliminary |
| 2014/09/08 | 7     | 1120          | 700         | P12        | Civil      | 1.2.300 | Yes        | Preliminary |
| 2014/09/09 | 1     | 160           | 100         | P12        | Civil      | 1.2.300 | Yes        | Preliminary |
| 2014/09/15 | 3     | 390           | 240         | P3         | Structural | 1.2.400 | Yes        | Site        |
| ...     | ...   | ...           | ...         | ...        | ...        |         | ...        | ...         |\
\\
\

@@

These three sources needed to be compiled into a single dataset that detailed one project per row as the model was to predict the overall profitability of a project. Before compilation could begin however, thorough cleaning was required where the data was plotted and statistically assessed so that outliers could be visually or statistically discovered. Outliers were then investigated for data entry errors. Many such errors were encountered. For example, a common glitch in the data collection process translated a user entering a 123 kilometer drive in a car as 123 hours. Or an expense printing claim of $9.50 would be translated to 9.5 hours spent on that job. Once these errors were detected, they were discussed with the company directors and appropriate corrective action to the data was taken. Mistakes in data entry were often able to be corrected by notes that a user had entered into the system. It is possible that some data entry errors were not detected and this is a problem that also must be dealt with during implementation of the model.

@@

Although dozens of variables were available from the initial dataset, further descriptive variables were engineered to trial in the models. This was particularly advantageous in the invoicing dataset and timesheet dataset where a project could have thousands of rows of relevant data that needed to be converted into a single row per project. All variables would eventually be tested for variable importance and predictive power with respect to project profitability. Examples of these engineered variables include:

@@

* Percent of hours performed by each professional role over the course of a project
* Timespan of the entered project hours
* Percent of hours performed by 'professional' employees as opposed to 'technical' employees
* Position of the employee that completed the most hours on each project
* Percent of hours done by the majority contributor to a project
* Total cost of employee hours per project
* Total cost of external subcontractors or disbursements per project
* Total number of employees that entered hours on each project
* Mean number of hours per day entered on a project
* Number of disciplines active in a project 
\ 
\

@@

* Total amount invoiced and remunerated per project
* Mean invoice size per project (for example if a project had four invoices totaling \$10,000, the mean invoice size would be \$2,500)over all invoices sent to a client accross all their projects. This gives an indication of whether the client generally does fast paced big jobs which would result in large monthly invoices for example or small jobs with small monthly invoiced amounts.
* Client invoice frequency which gives an indication of how much work the company does for that client. 
\ 
\

@@

* text analysis of project descriptions detected a list of key words that could classify projects into 16 categories. This key word analysis was done in conjunction with a company employee and the resulting classifications were reviewed by the employee to ensure accuracy. For example, project descriptions containing the words 'residence', 'house', 'apartment', 'home', 'townhouse', 'unit', and 'dwelling' were classified as 'residential'
* number of projects completed with each client and client contact

With the engineered variables, all three data sources could be represented in the single-row-per-project format. The three datasets were then combined and further variables were engineered using combinations of variables across the original data sets:

@@

Besides the engineered variables, original variables included in the project summary dataset were client industry, internal company discipline, job description, and post code. Once numerous potentially important variables were engineered, the most important ones could then be selected. This practice improves the accuracy of a model because unnecessary or irrelevant variables add noise to the prediction of target values. Variable selection also increases computational efficiency and simplifies understanding of the prediction structure [@Weisberg2005].

@@

After variable engineering, the contribution of each variable to the predictive model was assessed. To evaluate variable importance, several models predicting 'return per dollar' were built using all explanatory variables. The models then output which variables contributed significantly or most improved results. Different types of predictive models have different methods of calculating variable importance and since the best predictive model was unknown, a few importance outputs were reviewed. These included one linear and two ensemble tree methods. However, before variable importance models were built, outlier cases were deleted because they are often the result of unusual events or errors in data entry.

@@

Case study projects with extreme 'return per dollar' values were likely the result of special scenarios which the model is not intended to predict. For example, many projects appeared wildly profitable but on closer inspection these were all very small jobs. After discussion with the case study company, it was concluded that for those outliers, the small jobs most likely required so little time that busy employees did not bother writing down their hours for that job. Outliers were initially defined as 'return per dollar' values sitting outside 1.5 x the interquartile range (IQR) from the upper and lower quartiles. This rule was initially practised by Tukey, the inventor of the box plot [@Tukey1977]. All outlier projects determined by this method were reviewed with an employee from the case study company and based on their domain knowledge and assessment of the outlier cases, reasonable cut-off values for the range of return per dollar values were determined. In summary, projects with 'return per dollar' values greater than 3 were removed and one project with a very low 'return per dollar' was removed because it contained incorrect invoiced amounts.

@@

The three methods were anticipated to output different levels of influence for each variable. The tested methods were ANOVA Linear Regression, conditional inference forests (cForests), and Random Forests. It was expected that Random Forests would favour variables that had more categories, a well-known bias; which is also a bias that cForests have overcome [@Strobl2007]. ANOVA is based on linear theory which contrasts conditional forest and Random Forest's more free-form structure. Therefore, it is beneficial to compare the methods' assessments of variable importance.

When running variable importance tests, the dependent variable must be clearly defined. For this problem, the response variable could be continuous or categorical i.e. *how* profitable a project was (continuous) or whether a project was profitable or not (binary classification). The continuous response variable, 'reurn per dollar', was chosen for the variable importance models as the *degree* of profitability also gives the binary classification (profitable or unprofitable).

#### ANOVA Variable Importance

ANOVA is closely related to linear fit models but incorporates the analysis of differences in group means (i.e. categorical variables) [@Lunney1970]. Since the method assumes variables are normally distributed, numeric variables must be normalised before analysis to achieve a statistical distribution as close to Gaussian as possible. Therefore a disadvantage of this method is that linear relationships and normally distributed variables do not necessarily represent real world data accurately [@Breiman2001b].

In order to compare variable importance, two ANOVA outputs were assessed: the linear coefficients for each variable and the p-values for the coefficients [@Markham2016]. The coefficient does not indicate importance relative to the other variables because its value reflects values within the variable. The sign of the coefficient however indicates whether the variable is directly or inversely related to the response variable. A significance test for whether the coefficient equals zero is performed on each coefficient and the p-value is reported. Therefore, if the null hypothesis is rejected (p value < 0.05), changes in the explanatory variable are indeed related to changes in the response variable. P-values are a good indication of whether variables have a relationship with the target variable, however they do not necessarily rank importance. Precision of variable values can influence p-values, for example a variable with high precision will have a smaller p-value, whereas a variable measured roughly will have a higher p-value. This does not necessarily make the more precisely measured variable more important than the other. It is also important to keep in mind that, with a dozen variables, chance alone can produce a variable with a p-value < 0.001 7% of the time [@Rice1989].

@@

#### Random Forest Variable Importance

The Random Forest algorithm can produce a permutation variable importance for each variable as part of its output. Variable importance is represented as a score, which is calculated by performing permutations on each variable in the test data for each tree, then re-running the test data through the trees. More specifically, consider a single variable. The values of this variable in the out-of-bag sample (the test data that the Random Forest algorithm partitions before building each tree) are randomly permuted (reordered) for each tree. The purpose of permuting is to mimic the absence of that variable. The trees in the Random Forest are then run again. Variable importance can now be derived by comparing the results of the forest where the covariate has been permuted against the original Random Forest. If the accuracy of the permuted forest is much less than the original forest, then that variable was important. Permutation importance scores are calculated as the mean decrease in accuracy over all trees for the permuted variable in the model [@Breiman2005].

These scores provide valuable insight because they represent the impact of each predictor variable individually as well as in multivariate interactions with other predictor variables [@Strobl2007]. However, a severe disadvantage is that the scores are not reliable when the variables differ in their numeric scales or their counts of categories. Also, the importance of correlated predictors is overestimated and the algorithm tends to favour variables that have many possible splits or many missing values [@Strobl2007]. The intuition behind this is that if a variable has more points to split the data, there are more opportunities for it to split the response variable favourably. However this does not necessarily indicate the variable impacts the response variable the most. Overall, the permutation importance is useful to compare variables in a non-linear environment and can include multivariate interactions, but overestimates the importance of variables with with many categories and some numeric variables.

#### Cforest Variable Importance

Cforests are an alternative ensemble tree method that overcomes the aforementioned shortfalls of Random Forest scores. It is built from cTrees which are decision trees based on a conditional inference framework. The key difference between cTrees and standard decision trees is that a significance test is used for splitting instead of a purity measure such as the Gini coefficient (which is centered around information gain).

The procedure behind the significance test at each split in a cTree is as follows. To determine which explanatory variable should be used at each split, each variable is permuted in every possible way, and a correlation value is calculated between the tested variable and response variable, for each permutation. The unchanged variable correlation is then compared with the correlation values for all permutations of that variable. From this, a p-value for the true correlation value compared to the permuted correlation values can be calculated. The predictor variable with the lowest p-value is then selected as the splitting variable [@Hothorn2006]. Using an ensemble forest of these trees, variable importance is then calculated in the same fashion as Random Forests, via permuting a variable, re-running the forest and comparing the decrease in response variable accuracy. Several sources recommend cForests variable importance methods because its statistical p-value tests at each split removes the bias present in other tree ensemble variable importance outcomes such as Random Forests (as discussed) and Boosted Trees [@Strobl2007][@Strobl2009][@Hothorn2006].

@@

Limiting the predictive model to a concise set of meaningful variables reduces noise and improves predictions. Less variables and a simpler model is easier for stakeholders to understand [@Weisberg2005]. For these reasons, a subset of the most important variables were chosen before modelling. This was done by first eliminating outliers, then comparing important variables from ANOVA as well as Random Forests and cForests. It can be inferred from the literature that cForests would suit this project's data best as it is unbiased and not limited to linear theory [@Strobl2007].

@@

Previous studies predicting profitability in software and construction projects were predominantly limited to case-based reasoning, regression and Neural Networks. There is a gap in testing other sophisticated machine learning algorithms that are as powerful as Neural Networks but provide insight into the reasoning behind predictions. Ensemble trees (Boosted Trees and Random Forests) and Bayesian Networks fit these criteria. Regression and Naive Bayes models were also included as simple baseline models. Complex models should be measured against simple models that can be built at a fraction of the computational cost. The complete list of models tested in this study is:

* Regression - baseline model
* Naive Bayes - baseline model
* Bayesian Network
* Random Forest
* Gradient Boosted Trees

Some models have limitations and required additional data processing steps such as normalising numeric variables (Bayesian Networks and Regression) and discretising continuous variables (Bayesian Networks). Discretisation was performed by generating a hierarchical dendrogram of each variable to visualise the clusters. Between four and six clusters were chosen and summarised to find the maximum and minimum values within each cluster. For example, the diagram below illustrates the hierarchical dendrogram for timespan with 6 clusters boxed.

@@

The number of clusters was determined both visually from the dendrogram and experimentally. Visually, the height of the dendrogram 'branches' indicates the change in 'tightness' of the data points to their cluster centroids as the number of clusters increases. Tightness in the case of Ward's method is the sum of squared distances of each data point to the centroid of its respective cluster (ESS). For example, the height of the top horizontal bar indicates the ESS for one cluster and the height of the second from top horizontal bar is the ESS for two clusters. The difference in the two heights indicates the improvement in ESS by adding a cluster [@Putler2012]. As more clusters are added, decreases in ESS are less significant and the aim is to choose the number of clusters at an optimal level in this process.  After initial comparisons of predictive methods had been trialed, the number of clusters for each discretised variable was re-assessed by trialing one more or one less cluster and comparing model output.

@@

Although five methods were being tested, the response variable (profitability) could be numeric or categorical, i.e.

* the 'return per dollar' of each project (numerical) or 
* whether each project made a 'profit' or 'loss' (categorical)

This means the problem could be framed as either a regression or binary classification problem, and each predictive method could be applied to both. Since it was uncertain which framework would provide the most benefit, the models were built in both.

**you are here**

To compare the models, the root mean squared error (RMSE) statistic was used for regression models and the area under the receiver operating characteristic (ROC) curve (AUC) statistic was used to measure binary classification.

An ROC graph visualises the curve from which an AUC score is calculated. Its two axes are the false positive rate (FPR) on the x-axis and true positive rate (TPR) on the y-axis. In this problem a 'positive' is a loss-making job so a TP is a case when the model predicts a job will be loss making, and it did indeed lose money. A false positive (FP) is when the model predicts a job will be loss making but in reality it was profitable. These values are calculated as follows:

Only predictions made on the test data should be entered into the equations above, therefore a model must be first created using a training set. The model can then make predictions on the test set. Binary class model predictions are probability values between 0 and 1, which cannot be entered into the equations above. In order to classify each case in the test set as 0 or 1, a probability threshold must be chosen. That is, if a probability threshold is chosen to be 0.6, each case in the test set can then be classified as 0 or 1 depending on whether its predicted probability is greater or less than 0.6. Once the 0 or 1 classifications have been assigned for a certain threshold, the TPR and FPR can be calculated for that threshold. This is plotted as a single point on the ROC curve, say point C. Refer the diagram below:

A model that is perfectly classified would have all positives correctly classified (1.0 True Positive Rate) and no incorrectly classified positives (0.0 False Positive Rate). If a model has a 1.0 TPR and 1.0 FPR (top right corner) it has correctly classified all positives at the expense of incorrectly classifying all negatives as positive. It is 'dumbly' classifying all cases as positive. If a model classifies 0.8 of its positives as TP but also 0.8 of its negative cases as positive it is 'dumb' in a similar way to the previous example. There is an 80% chance *any* case will be classified as positive. Therefore, a point that lies on the diagonal line classifies cases with the same capability as random chance.

To plot the curve in an ROC, a series of points and their coordinates must first be calculated. This is done by choosing numerous threshold probability values between 0 and 1, then classifying each probability outcome in the test set as 0 or 1 based on the threshold value. Finally the TPR and FPR for that threshold point can be calculated which are the coordinates. Once enough points are plotted, a curve can be drawn. The closer the curve reaches toward the top left hand corner, the closer the algorithm is to perfectly predicting positive and negative cases at some optimal threshold. A curve that squarely reaches the top left hand corner would have an area under its curve (AUC) of 1. An AUC between 0.5 and 1 means the model is performing better than random chance (since the diagonal line represents a model randomly assigning positives to all cases at a certain rate and would have an AUC of 0.5).

For binary classification, AUC is a more useful statistic than classification accuracy when output is available as a probability which can be applied to the problem. This is pertinent when it is useful to rank the outcomes and trial probability thresholds that will optimise the rates of true positives and true negatives [@Huang2005]. Note the 'probability' outputs from the models are not true statistical probabilities, but a score the model has assigned to a case using its own measure of certainty.

In order to compare which models performed significantly better than others, an adequate sample size of results statistics is required. Multiple models could be made by using different data in the training vs. testing sets, each providing a resulting test statistic (RMSE or AUC). 5 fold cross validation was used, which created 5 models, then the division of 5 folds was repeated using a different random seed to create another set of 5 training/testing data sets. Initially, 20 models of each method were created in this fashion. Then a two-sample power calculation was run using the two sets of 20 results to determine the sample size to achieve a statistical power of 0.8. Once the number of required samples was determined, more models were built via 5 fold cross validation to obtain additional results statistics. The highest performing models were advanced to the next stage where various methods of combining, or blending, methods were tested.

@@

All methods except gradient Boosted Trees and Naive Bayes could not handle missing data. Therefore, preliminary runs of each method used subsets of the data which had complete data. It was possible that if missing data was imputed, predictions from the models that were limited by missing data could improve. Also, a complete data set allows for complete sets of predictions from each method, and these predictions could then be blended to further improve results. The procedure behind model blending is addressed further in the next section. The imputed data set was compared to gradient Boosted Tree results since Boosted Trees can handle missing values and should perform equally well with imputed data.

The MICE Random Forest method was chosen for imputation because it has been proven to work well with complex data sets [@Shah2014]. It first performs a standard Random Forest imputation of missing values which are treated as initial 'place holders'. Random forest imputation is a complex process which works by first doing a rough imputation of the missing values. Then a ten-tree forest is run to calculate proximities between all cases. Proximities between two cases represent the number of trees in a forest where the two cases were assigned to the same end node in a tree. Once proximities are calculated, missing values for a variable (say variable 'm') are filled by an average of the other complete 'm' values weighted by the proximity of the case representing each 'm' value.

@@
